---
title: "Characterization of the CRS dampened genes"
author: "Pierre-Luc Germain"
date: "2023-07-05"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(dev="CairoPNG")
suppressPackageStartupMessages({
  library(sechm)
  library(edgeR)
  library(stageR)
  library(ggplot2)
})
theme_set(theme_classic())
source("../functions/functions.R")
source("../functions/clusterProfiling.R")
th.lfc <- log2(1.2)
```

# Overall

```{r}
se <- readRDS("all.full.SE.rds")
d <- readRDS("damping.rds")
dt <- split(d,d$TimePoint)
for(f in names(dt)){
  g <- intersect(getDEGs(se, f, lfc.th=th.lfc), row.names(se))
  row.names(dt[[f]]) <- dt[[f]]$gene
  dt[[f]] <- dt[[f]][g,]
  dt[[f]]$category <- cut(dt[[f]]$damping, breaks=c(0,0.2,0.75,1), labels=c("low","medium","high"), include.lowest = TRUE)
  rowData(se)[[paste0("damping",f)]] <- NA_real_
  rowData(se)[row.names(dt[[f]]),paste0("damping",f)] <- dt[[f]]$damping
  dt[[f]] <- dt[[f]][order(sign(dt[[f]]$logFC)*dt[[f]]$damping),]
}
```

```{r}
# take the most significant damping per gene
ag1 <- d[!duplicated(d$gene),]
row.names(ag1) <- ag1$gene
ggplot(ag1[which(ag1$FDR<0.05 & abs(ag1$logFC)>th.lfc),], aes(damping, logFC)) + 
  geom_hline(yintercept=0, linetype="dashed") + ggpointdensity::geom_pointdensity() +
  scale_colour_viridis_c(guide = "none") + labs(x="Damping coefficient", y="logFC upon stress")
```

There's no relationship between the degree of damping and the strength of the response to stress.

```{r}
ag1 <- ag1[,c("gene","TimePoint","logFC","logCPM","FDR","logFC.CRS","PValue.CRS","FDR.CRS","damping")]
names(ag1)[ncol(ag1)] <- "topDamping"
# take the weighted mean of the damping
ag2 <- sapply(split(d[,c("damping","PValue.CRS")], d$gene), FUN=function(x){
  weighted.mean(x$damping, pmax(0,-log10(x$PValue.CRS)-0.3),na.rm=TRUE)
})
ag1$wmDamping <- ag2[row.names(ag1)]
ag <- setNames(ag1$wmDamping, ag1$gene)
ag <- ag[names(ag) %in% row.names(se)]
alldegs <- union(row.names(se)[which(rowData(se)$FDR<0.05)],
                 length(unique(unlist(getDEGs(se, lfc.th=th.lfc)))))
adegs2 <- intersect(alldegs, names(ag))
ag <- ag[adegs2]
ag <- ag[!is.na(ag)]
rowData(se)$maxDamping <- NA_real_
rowData(se)[names(ag),"maxDamping"] <- as.numeric(ag)

# # find the elbow
# elbow <- ag[which.max(sqrt((rank(ag)/length(ag))^2+(1-ag)^2))]
# mm <- mixtools::normalmixEM(ag, k=3)
# dc <- c("low","medium","high")[apply(mm$posterior,1,which.max)]
dc <- cut(ag, breaks=c(0,0.1,0.5,0.8,1), labels=c("none","low","medium","high"), include.lowest = TRUE)

dampingPlot(data.frame(gene=names(ag), rank=rank(ag)/length(ag), damping=ag, category=dc),
            goi=c("Adat3", "Apold1", "Npas4", "Fosb", "Fos", "Sgk1", "Arc", "Junb", "Dio2","Cwc25","Creb5"))

rowData(se)$dampingCategory <- NA
rowData(se)[names(ag),"dampingCategory"] <- as.character(dc)
rowData(se)$dampingCategory <- factor(rowData(se)$dampingCategory,levels(dc))
saveRDS(se, file="all.full.SE.rds")
```

## Damping patterns


```{r, eval=FALSE}
degs <- unique(unlist(lapply(getDEA(se)[1:3], FUN=getDEGs, lfc.th = th.lfc, fdr.th=0.01)))
a <- assays(se)$scaledLFC[intersect(degs,row.names(se)),se$TimePoint %in% c("45min","1h30","3h")]
w <- which(abs(a)>5)
a[w] <- 5*sign(a[w])
set.seed(42)
res <- Spectrum::Spectrum(t(a), runrange=TRUE, krangemax = 10)
cl <- split(row.names(a), res[[10]]$assignments)
saveRDS(cl, file="clustering_res.rds")
```


```{r}
cl <- readRDS("clustering_res.rds")
sechm(se, cl, gaps_at="TimePoint", assayName = "scaledLFC", row_title_gp=gpar(fontsize=12), column_title_gp=gpar(fontsize=12))
```

```{r}
gl <- list(
  "blunted\nat 45min"=cl[["3"]],
  "faster return\nto baseline"=unlist(cl[c("1","4")], use.names = FALSE),
  "ablated\nat 3h"=cl[["8"]]
  
)
saveRDS(gl, "full.dampingClusters.rds")
gl.hm <- sechm(se, gl, gaps_at="TimePoint", assayName = "scaledLFC", top_annotation=c("TimePoint","Treatment"),
               row_title_gp=gpar(fontsize=12), column_title_gp=gpar(fontsize=12),
               row_title_rot=0, use_raster=TRUE)
gl.hm
```

Enrichment vs all expressed genes:

```{r}
gsets <- getMsigSets()
WP <- gsets[grep(":WIKIPATHWAYS",names(gsets))]
regs <- readRDS("../extDat/collecTRI.regulons.rds")
regs <- split(tools::toTitleCase(tolower(regs$target)), regs$source)
expressed_genes <- row.names(se)[which(!is.na(rowData(se)$FDR))]
sigs <- lapply(gl, FUN=function(x){
  setNames(expressed_genes %in% x, expressed_genes)
})
gl.reg.vsall <- lapply(sigs, sets=regs, FUN=enrichmentTests)
gl.wp.vsall <- lapply(sigs, sets=WP, FUN=enrichmentTests)
gl.ora.vsall <- lapply(sigs, sets=gsets, FUN=enrichmentTests, alternative = "two.sided")

sigTerms <- unique(unlist(lapply(gl.ora.vsall, FUN=function(x){
  x <- x[x$FDR<0.05 & x$overlap>=5,]
  if(sum(x$enrichment>3)<2) return(head(row.names(x)))
  x <- x[x$enrichment>3,]
  return(head(row.names(x)))
})))
ora <- dplyr::bind_rows(lapply(gl.ora.vsall, FUN=function(x){
  x <- x[sigTerms,]
  x$pathway <- sigTerms
  x
}), .id="Cluster")
ora$pathway <- breakStrings(gsub("_"," ",gsub("^[A-Z]*:.*:","",ora$pathway)))
ora$Cluster <- factor(ora$Cluster, names(gl)[c(3,2,1)])


gl.enrP <- ggplot(ora, aes(Cluster, pathway, size=-log10(pvalue), fill=enrichment, colour=FDR<0.05)) + 
    geom_point(pch=21) + scale_fill_gradient2() + scale_size_continuous(range=c(2,8)) + 
  scale_colour_manual(values=c("FALSE"="white", "TRUE"="black")) + 
  theme_bw() + theme(axis.text.x=element_text(angle=90, hjust=1,vjust=0.5),
                       axis.title=element_blank()) + labs(fill="fold-\nenrichment")

gl.enrP
```

Faster return to baseline:
  GO:BP:GOBP_RESPONSE_TO_ABIOTIC_STIMULUS (2-fold, FDR~0.00014) and GOBP_CELLULAR_RESPONSE_TO_CORTICOSTEROID_STIMULUS (3.4-fold, FDR~0.014)
Top CollectTRI enrichment: NR3C1 (3-fold, FDR 1e-7), followed by PPARG and HIF1A.

Ablated at 3h: no enrichment

Blunted at 45min: 
orexin & corticotropin, vasculature development
CREB1 (2.3-fold, FDR~0.00054)...


vs response genes:

```{r}
degs <- unique(unlist(lapply(getDEA(se)[1:5], FUN=getDEGs, lfc.th = th.lfc)))
sigs <- lapply(gl, FUN=function(x){
  setNames(degs %in% x, degs)
})
gl.reg <- lapply(sigs, sets=regs, FUN=enrichmentTests)
gl.wp <- lapply(sigs, sets=WP, FUN=enrichmentTests)
gl.ora <- lapply(sigs, sets=gsets, FUN=enrichmentTests, alternative = "two.sided")
sigTerms <- unique(unlist(lapply(gl.ora, FUN=function(x) head(row.names(x)[which(x$FDR<0.25)]))))
sigTerms <- c(sigTerms, "GO:BP:GOBP_CELLULAR_RESPONSE_TO_TOXIC_SUBSTANCE") # to have one top per cluster
ora <- dplyr::bind_rows(lapply(gl.ora, FUN=function(x){
  x <- x[sigTerms,]
  x$pathway <- sigTerms
  x
}), .id="Cluster")
ora$pathway <- breakStrings(gsub("_"," ",gsub("^[A-Z]*:.*:","",ora$pathway)))
ora$Cluster <- factor(ora$Cluster, names(gl)[c(3,2,1)])


gl.enrP2 <- ggplot(ora, aes(Cluster, pathway, size=-log10(pvalue), colour=enrichment)) + 
  geom_point() + scale_colour_gradient2() + scale_size_continuous(range=c(2,8)) + 
  theme_bw() + theme(axis.text.x=element_text(angle=90, hjust=1,vjust=0.5),
                     axis.title=element_blank())
gl.enrP2
```

```{r}
enr <- list(
  "vsExpressed"=list(regulons=gl.reg.vsall, WP=gl.wp.vsall, ORA=gl.ora.vsall),
  "vsResponse"=list(regulons=gl.reg, WP=gl.wp, ORA=gl.ora)
)
saveRDS(enr, "dampingClusters_enrichments.rds")
```

```{r}
gl2 <- list(
  "(down) not damped"=cl[["5"]],
  "(down) faster return\nto baseline"=cl[["9"]],
  "(down) ablated\nat 3h"=c(cl[["6"]],cl[["7"]])
)
sigs <- lapply(gl2, FUN=function(x){
  setNames(expressed_genes %in% x, expressed_genes)
})
gl.reg.vsall <- lapply(sigs, sets=regs, FUN=enrichmentTests)
gl.wp.vsall <- lapply(sigs, sets=WP, FUN=enrichmentTests)
gl.ora.vsall <- lapply(sigs, sets=gsets, FUN=enrichmentTests, alternative = "two.sided")

sigTerms <- unique(unlist(lapply(gl.ora.vsall, FUN=function(x){
  x <- x[x$FDR<0.05 & x$overlap>=5,]
  if(sum(x$enrichment>3)<2) return(head(row.names(x)))
  x <- x[x$enrichment>3,]
  return(head(row.names(x)))
})))
ora <- dplyr::bind_rows(lapply(gl.ora.vsall, FUN=function(x){
  x <- x[sigTerms,]
  x$pathway <- sigTerms
  x
}), .id="Cluster")
ora$pathway <- breakStrings(gsub("_"," ",gsub("^[A-Z]*:.*:","",ora$pathway)))
ora$Cluster <- factor(ora$Cluster, names(gl2)[c(3,2,1)])

saveRDS(ora, file="dampingClusters_enrichments.down.rds")

gl.enrP <- ggplot(ora, aes(Cluster, pathway, size=-log10(pvalue), fill=enrichment, colour=FDR<0.05)) + 
    geom_point(pch=21) + scale_fill_gradient2() + scale_size_continuous(range=c(2,8)) + 
  scale_colour_manual(values=c("FALSE"="white", "TRUE"="black")) + 
  theme_bw() + theme(axis.text.x=element_text(angle=90, hjust=1,vjust=0.5),
                       axis.title=element_blank()) + labs(fill="fold-\nenrichment")
```

